### 2016-02-25 discussion.Rmd
Final submission.

Two major meta-analyses of WMHs have been published covering the periods of 1966-2009 and 2010-2015, respectively. The earlier meta-analysis covered 53 longitudinal studies that included samples of high-risk populations, i.e., patients selected for a specific disease or condition such as hypertension, whereas other studies recruited samples of the general population. Longitudinal studies of samples representative of the general population are more relevant to the focus of the present paper. Debette & Markus [@Debette:2010aa] found that the presence of WMHs was related to subsequent cognitive decline, a higher risk of developing dementia, stroke, and of mortality. Lesion volume at baseline was also predictive of cognitive decline. Limitations of this meta-analysis include heterogeneity in the method of measuring WMHs; some studies used automated volumetric measurement, whereas others used a visual rating scale. The studies analyzed by Debette & Markus were limited to the occurrence of one of the aforementioned conditions which they analyzed by hazard ratios.

\replaced[id=dt]{The more}{A} recent meta-analysis by Kloppenborg et al. [@Kloppenborg:2014aa] of 23 cross-sectional studies reporting MRI and concurrent neuropsychological results in patients with heterogeneous diagnoses but without previously diagnosed cognitive impairment, found that WMHs were associated with cognitive deficit (effect size of -0.10, 95% CI: -0.13 to -0.08) after controlling for age. These studies also differed in the metric used to measure the WMHs, including volume, % of total intracranial volume, and a visual rating score. The effect size for the association with cognitive deficit in these cross-sectional studies did not differ significantly across various cognitive domains or the method of measuring lesion volume.  Among eight longitudinal studies analyzed by Kloppenborg et al that included a follow-up MRI and also controlled for age, the effect size for the association of progression in WMHs and cognitive impairment was -0.16 (95% CI:-0.27 to -0.09). This association was stronger for attention and executive function than for memory and processing speed. Although baseline WMHs were predictive of cognitive deficit at follow-up in the seven studies which did not repeat MRI, the effect size was smaller [-0.10 (95% CI: 0.13 to -0.05) than in the longitudinal studies that calculated progression in WMHs. In summary, progression of WMHs seen on repeat MRI has a stronger relation to cognitive deficit than concurrent imaging findings. These meta-analyses support the rationale for repeating an MRI in patients younger than 50 years whose initial scan shows WMHs.
Despite the above-described associations between WMHs, cognitive decline, increased risk of developing dementia, and mortality, these lesions receive little attention in current clinical workflows. When reported in a standard neuroradiologist interpretation, they are typically handled as incidental findings and are assigned little clinical significance. This likely reflects the impracticality of performing a detailed assessment of number, volume, and distribution within a qualitative neuroradiologist interpretation as well as the lack of correlative information on how the presence and distribution of these lesions may inform a diagnosis and prognosis in the appropriate clinical setting. To date, automated or semi-automated tools for the detection of WMHs have lacked the specificity and efficiency for the mining of large-scale datasets to generate highly granular data on whether these lesions possess any true diagnostic or prognostic value in the setting of a specific disease process. The present communication describes a supervised statistical learning tool that is appropriate for the application to such large-scale datasets.

@@

<!--
\listofchanges

\clearpage
-->

# References

>>>>

Two major meta-analyses of WMHs have been published covering the periods
prior to 2010 [@Debette:2010aa] and after [@Kloppenborg:2014aa]. The earlier meta-analysis
covered 53 longitudinal studies that included samples of high-risk
populations, i.e., patients selected for a specific disease or condition
such as hypertension, whereas other studies recruited samples of the
general population. Longitudinal studies of samples representative of
the general population are more relevant to the focus of the present
paper. Debette & Markus [@Debette:2010aa] found that the presence of
WMHs was related to subsequent cognitive decline, a higher risk of
developing dementia, stroke, and of mortality. Lesion volume at baseline
was also predictive of cognitive decline. Limitations of this
meta-analysis include heterogeneity in the method of measuring WMHs;
some studies used automated volumetric measurement, whereas others used
a visual rating scale. The studies analyzed by Debette & Markus were
limited to the occurrence of one of the aforementioned conditions which
they analyzed by hazard ratios.

The more recent meta-analysis by Kloppenborg et al.
[@Kloppenborg:2014aa] of 23 cross-sectional studies reporting MRI and
concurrent neuropsychological results in patients with heterogeneous
diagnoses but without previously diagnosed cognitive impairment, found
that WMHs were associated with cognitive deficit (effect size of -0.10,
95% CI: -0.13 to -0.08) after controlling for age. These studies also
differed in the metric used to measure the WMHs, including volume, % of
total intracranial volume, and a visual rating score. The effect size
for the association with cognitive deficit in these cross-sectional
studies did not differ significantly across various cognitive domains or
the method of measuring lesion volume.  Among eight longitudinal studies
analyzed by Kloppenborg et al that included a follow-up MRI and also
controlled for age, the effect size for the association of progression
in WMHs and cognitive impairment was -0.16 (95% CI:-0.27 to -0.09). This
association was stronger for attention and executive function than for
memory and processing speed. Although baseline WMHs were predictive of
cognitive deficit at follow-up in the seven studies which did not repeat
MRI, the effect size was smaller [-0.10 (95% CI: 0.13 to -0.05) than in
the longitudinal studies that calculated progression in WMHs. In
summary, progression of WMHs seen on repeat MRI has a stronger relation
to cognitive deficit than concurrent imaging findings. These
meta-analyses support the rationale for repeating an MRI in patients
younger than 50 years whose initial scan shows WMHs.

Despite the
above-described associations between WMHs, cognitive decline, increased
risk of developing dementia, and mortality, these lesions receive little
attention in current clinical workflows. When reported in a standard
neuroradiologist interpretation, they are typically handled as
incidental findings and are assigned little clinical significance. This
likely reflects the impracticality of performing a detailed assessment
of number, volume, and distribution within a qualitative
neuroradiologist interpretation as well as the lack of correlative
information on how the presence and distribution of these lesions may
inform a diagnosis and prognosis in the appropriate clinical setting. To
date, automated or semi-automated tools for the detection of WMHs have
lacked the specificity and efficiency for the mining of large-scale
datasets to generate highly granular data on whether these lesions
possess any true diagnostic or prognostic value in the setting of a
specific disease process. The present communication describes a
supervised statistical learning tool that is appropriate for the
application to such large-scale datasets.

@@

### 2016-02-25 format.Rmd
Final submission.

---
output:
  word_document:
    fig_caption: true
  pdf_document:
    fig_caption: true
    latex_engine: xelatex
    keep_tex: yes
header-includes:
   - \usepackage{booktabs}
   - \usepackage[final]{changes}
   - \usepackage[font={small},labelfont=bf,labelsep=colon]{caption}
   - \linespread{1.5}
   - \usepackage[compact]{titlesec}
   - \usepackage{enumitem}
   - \usepackage{tikz}
   - \def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}
   - \setlist{nolistsep}
   - \titlespacing{\section}{2pt}{*0}{*0}
   - \titlespacing{\subsection}{2pt}{*0}{*0}
   - \titlespacing{\subsubsection}{2pt}{*0}{*0}
   - \setlength{\parskip}{3pt}
   - \setremarkmarkup{(#2)}
   - \definechangesauthor[name={Nick Tustison}, color=red]{nt}
   - \definechangesauthor[name={James Stone}, color=magenta]{js}
   - \definechangesauthor[name={Lisa Wilde}, color=cyan]{lw}
   - \definechangesauthor[name={Andy Mayer}, color=green]{am}
   - \definechangesauthor[name={Harvey Levin}, color=blue]{hl}
   - \definechangesauthor[name={Brian Taylor}, color=purple]{bt}
   - \definechangesauthor[name={David Tate}, color=orange]{dt}
bibliography: references.bib
csl: council-of-science-editors.csl
fontsize: 11pt
mainfont: Georgia
geometry: margin=1.0in
---

>>>>

---
output:
  word_document:
    fig_caption: true
  pdf_document:
    fig_caption: true
    latex_engine: xelatex
    keep_tex: yes
header-includes:
   - \usepackage{booktabs}
   - \usepackage[final]{changes}
   - \usepackage[font={small},labelfont=bf,labelsep=colon]{caption}
   - \linespread{2}
   - \usepackage[compact]{titlesec}
   - \usepackage{enumitem}
   - \usepackage{tikz}
   - \def\checkmark{\tikz\fill[scale=0.4](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}
   - \setlist{nolistsep}
   - \titlespacing{\section}{2pt}{*0}{*0}
   - \titlespacing{\subsection}{2pt}{*0}{*0}
   - \titlespacing{\subsubsection}{2pt}{*0}{*0}
   - \setlength{\parskip}{3pt}
   - \setremarkmarkup{(#2)}
   - \definechangesauthor[name={Nick Tustison}, color=red]{nt}
   - \definechangesauthor[name={James Stone}, color=magenta]{js}
   - \definechangesauthor[name={Lisa Wilde}, color=cyan]{lw}
   - \definechangesauthor[name={Andy Mayer}, color=green]{am}
   - \definechangesauthor[name={Harvey Levin}, color=blue]{hl}
   - \definechangesauthor[name={Brian Taylor}, color=purple]{bt}
   - \definechangesauthor[name={David Tate}, color=orange]{dt}
bibliography: references.bib
csl: council-of-science-editors.csl
fontsize: 11pt
mainfont: Georgia
geometry: margin=1.0in
---

### 2016-02-25 intro.Rmd
Final submission.

Machine learning and pattern recognition techniques have seen increased application
for various medical image analysis workflows (see, for example, the annual
Workshop on Machine Learning in Medical Imaging held in conjunction with the Medical
Image Computing and Computer-Aided Intervention (MICCAI) international meeting [@mlmi2015]).
Popular techniques such as support vector machines and neural networks have been applied
successfully to clinically relevant imaging tasks such as supervised image segmentation
(e.g., [@Bauer:2011aa]) and diagnostic prediction (e.g., [@Tong:2014aa;@Liu:2013aa]).
Facilitating the current employment of such techniques are the number of available imaging
data sets [@Van-Horn:2014aa] and the public availability of data science packages such as
SciPy [@scipy] and the R project for statistical computing [@R] and their associated
extensions.

The random forests framework [@breiman2001] is a popular machine learning technique that has demonstrated significant \replaced[id=dt]{utility}{utitility} for supervised segmentation tasks (e.g.,
normal human brain segmentation [@yi2009]) and other computer vision applications
(e.g., human gait detection [@viola2005]).
In the context of neuropathology, random forest-based paradigms have been employed in the
delineation of multiple sclerosis lesions [@geremia2011], stroke lesions [@Pustina:2016aa],
and brain tumors [@geremia2012;@bauer2012;@zikic2012;@Tustison:2015aa] for both
single and multi-modal acquisition protocols.  Of note, these
latter random forest approaches for brain tumor segmentation have performed well in recent
international competitions \replaced[id=dt]{established in response to the lack of objective comparisons between segmentation algorithms (i.e.,  the Multimodal Brain Tumor Segmentation (BRATS) challenge was initiated in 2012}{
In response to the lack of objective comparisons between segmentation algorithms, the Multimodal Brain Tumor Segmentation (BRATS) challenge was initiated in 2012 [@Menze:2015aa] and has continued every year since under the auspices of the MICCAI conference.}[@Menze:2015aa].

Random forests are conceptually straightforward [@breiman2001].  They consist of ensembles of decision trees
that are built from training data.  Once constructed, data to be classified is "pushed"
through each decision tree resulting in a single classification "vote" per tree.  These
votes are then used for regression or classification of the data.  Although decision
trees had been extensively studied \deleted[id=dt]{previously}, the success of employing collections of
such weak learners for boosting machine learning performance
(e.g., AdaBoost [@schapire1990;@freund1997]) influenced the similarly sytled conglomeration
of decision trees into "forests" with randomized node optimization [@ho1995;@amit1997].
Finally, Breiman [@breiman2001] improved accuracy by random sampling of \deleted[id=dt]{the} training
data (i.e., "bagging") resulting in the current random forest technique \added[id=dt]{applied here}.

In this work, we develop a concatenated random forest framework with a tailored contextual
feature image set (both spatial and intensity-based)
for segmenting \replaced[id=dt]{WMHs}{white matter hyperintensities} in \replaced[id=dt]{a arge TBI cohort}{traumatic brain injury cohorts}.  Additionally,
the entire framework is provided publicly through the well-known open-source
\added[id=lw]{Advanced Normalization Tools} (ANTs)[^i1] and ANTsR[^i2] toolkits.
\replaced[id=dt]{Further motivating this research is the availability of several large publicly available  imaging data sets that permits testing reproducibility of this automate routine for WMH segmentation and quantification.}{Further motivating this research of this work is
the public availability of the imaging data thus permitting full reproducibility
of the results reported and discussed.}

>>>>

Machine learning and pattern recognition techniques have seen increased application
for various medical image analysis workflows (see, for example, the annual
Workshop on Machine Learning in Medical Imaging held in conjunction with the Medical
Image Computing and Computer-Aided Intervention (MICCAI) international meeting).
Popular techniques such as support vector machines and neural networks have been applied
successfully to clinically relevant imaging tasks such as supervised image segmentation
(e.g., [@Bauer:2011aa]) and diagnostic prediction (e.g., [@Tong:2014aa;@Liu:2013aa]).
Facilitating the current employment of such techniques are the number of available imaging
data sets [@Van-Horn:2014aa] and the public availability of data science packages such as
SciPy [@scipy] and the R project for statistical computing [@R] and their associated
extensions.

The random forests framework [@breiman2001] is a popular machine learning technique that has demonstrated significant utility for supervised segmentation tasks (e.g.,
normal human brain segmentation [@yi2009]) and other computer vision applications
(e.g., human gait detection [@viola2005]).
In the context of neuropathology, random forest-based paradigms have been employed in the
delineation of multiple sclerosis lesions [@geremia2011], stroke lesions [@Pustina:2016aa],
and brain tumors [@geremia2012;@bauer2012;@zikic2012;@Tustison:2015aa] for both
single and multi-modal acquisition protocols.  Of note, these
latter random forest approaches for brain tumor segmentation have performed well in recent
international competitions established in response to the lack of objective comparisons between segmentation algorithms (i.e.,  the Multimodal Brain Tumor Segmentation (BRATS) challenge was initiated in 2012 [@Menze:2015aa].

Random forests are conceptually straightforward [@breiman2001].  They consist of ensembles of decision trees
that are built from training data.  Once constructed, data to be classified is "pushed"
through each decision tree resulting in a single classification "vote" per tree.  These
votes are then used for regression or classification of the data.  Although decision
trees had been extensively studied, the success of employing collections of
such weak learners for boosting machine learning performance
(e.g., AdaBoost [@schapire1990;@freund1997]) influenced the similarly sytled conglomeration
of decision trees into "forests" with randomized node optimization [@ho1995;@amit1997].
Finally, Breiman [@breiman2001] improved accuracy by random sampling of training
data (i.e., "bagging") resulting in the current random forest technique applied here.

In this work, we develop a concatenated random forest framework with a tailored contextual
feature image set (both spatial and intensity-based)
for segmenting WMHs in a large TBI cohort.  Additionally,
the entire framework is provided publicly through the well-known open-source
Advanced Normalization Tools (ANTs)[^i1] and ANTsR[^i2] toolkits.
Further motivating this research is the availability of several large publicly available  imaging data sets that permits testing reproducibility of this automated routine for WMH segmentation and quantification.

### 2016-02-25 methods.Rmd
Final submission.

MR images utilized for this initial report were acquired from 24 individuals scanned at a single site involved in the Chronic Effects of Neurotrauma Consortium’s (CENC) observational study (\added[id=nt,remark=NT: reference needed?]{see [Walker], this issue}). Briefly, participants were Operation Enduring Freedom/ Operation Iraqi Freedom /Operation New dawn (OEF/OIF/OND) era service members and veterans between the ages of 18--60 who have experienced combat situations and are between X and X (period of time) from deployment. \added[id=js,remark={JS: Data needed.}]{The feature images consisted of subjects aged $X.X \pm X$ years (range X – X years), XX percent male, with at least one or more potential concussive events (mean = X, SD = X, range X-X).}

@@

Crucial to these supervised segmentation approaches are the creation and selection of
"features" as input in conjunction with
\replaced[id=lw,remark=NT: I replaced ground-truth since it is problematic anyway]
{expertly identified structures of interest}{the ground-truth} for model construction.  For the targeted application in this work (i.e., WMHs),
regression/classification are performed at the voxelwise level.  In other words, each voxel
within the region of interest is sent through the ensemble of decision trees and receives
a set of classification votes from each tree thus permitting a regression or classification
solution.  Since this procedure is performed at the voxelwise level, intensity information
alone is insufficient for good segmentation performance due to the lack of spatial context.
For example, as pointed out in [@Neema:2009aa], higher intensities can be found at the
periventricular caps in normal subjects which often confounds automated lesion detection
algorithms.  Other potential confounds include MR signal inhomogeneity and noise.  Therefore,
even though machine learning and pattern recognition techniques are extremely powerful and
have significant potential, just as crucial to outcome is the creative construction and
deployment of salient feature images which we detail below.

@@

@@

@@

![Representation of Stage 1 feature images for subject 01C1019.  The
FLAIR, T1-, and T2-weighted images are \added[id=am,remark=NT: no need for 12 dof for
intra-subject normalization]{rigidly} pre-aligned
[@Avants:2014aa] to the space of the T1 image.  The three modality images are then preprocessed
(N4 bias correction [@Tustison:2010ac] and adaptive denoising [@Manjon:2010aa]) followed by
application of standard ANTs brain extraction and $n$-tissue segmentation protocols using
the MMRR symmetric template and corresponding priors [@Tustison:2014ab] applied to the T1 image.
The feature images are then generated for voxelwise input to the RF model which results in
the voting maps illustrated on the right.  This gives a probabilistic classification of
tissue type.  Not shown are the probability and voting images for the brain stem and
cerebellum.](Figures/featureImages.png)

As mentioned previously, input for each subject comprises FLAIR, T1-, and T2-weighted
acquisitions.  The FLAIR and T2 images are rigidly registered to the T1 image using
the open-source Advanced Normalization Tools (ANTs) [@Avants:2014aa].  The aligned images
are then preprocessed using the denoising algorithm of [@Manjon:2010aa] followed by
N4 bias correction [@Tustison:2010ac] which are then normalized to the intensity
range $[0,1]$.  Although we could have used an alternative intensity standardization
algorithm (e.g., [@nyul2000]), we found that a simple linear rescaling produced better results similar \added[id=dt]{to previous work}[@Tustison:2015aa].

@@

The T1 image is then processed via the ANTs brain
extraction and normal tissue segmentation (i.e., the pathology is discounted which is
a fairly safe assumption for the T1 image) protocols as part of the
cortical thickness pipeline described in [@Tustison:2014ab].
\added[id=hl]{This protocol involves preprocessing using N4 bias correction followed by a template-based strategy for brain extraction.  Once the
brain has been extracted, we apply a Bayesian-based segmentation algorithm
with a template-based prior probability strategy to segment the
parenchymal tissue types.  The result is} a mask delineating the brain parenchyma and probabilistic estimates of the
\replaced[id=dt]{CSF}{cerebrospinal fluid (csf)}, gray matter, white matter, deep gray matter, brain stem, and
cerebellum.  These provide the expertly annotated labels for the first six tissue labels
given above.  The \replaced[id=dt]{WMHs}{white matter hyperintensities} were manually identified by one of the authors
(J. R. S.) using the ITK-SNAP tool [@Yushkevich:2006aa].  Segmentation is performed using
the ANTs Atropos tool [@Avants:2011aa] and
multi-model optimal symmetric shape/intensity templates [@Tustison:2015aa] created from the public
MMRR data set  [@landman2011] (cf Figure 3).

@@

To model the intensity information the first set of images simply includes the
preprocessed and normalized intensity FLAIR, T1, and T2 image voxel values.  We also calculate a set
of neighborhood statistics (mean, standard deviation, and skewness) feature images using
a \added[id=am]{Manhattan} radius of one voxel \added[id=nt]{given the typical size
of individual WMHs}. For each of the normalized
images, we calculate the difference in intensities with the corresponding warped template
component.  Previous success in the international brain tumor segmentation
competition [@Menze:2015aa] was based on an important set of intensity features that were created
from multi-modal templates mentioned previously [@Tustison:2015aa] \added[id=nt]{and listed in Table 1}.  We employ the
same strategy here.  For example, the template
difference feature image for the FLAIR image, $S_{FLAIR}$ is calculated as:

@@

where $\phi_b: S  \leftrightarrow \underset{b}{\leftrightsquigarrow} T$ is the transform
which maps from the individual subject space to the
template space and $T_{FLAIR}$ is the FLAIR template component.  Also,
to take advantage of the \added[id=hl,remark=NT:  we are using the gross bilateral symmetry of the brain in an analogous sense that if we know the left eye location and color then we can accurately predict the location and color of the right eye even though a particular face might not be exactly symmetric]{gross} bilateral symmetry of the normal brain (in terms of both shape and intensity), and the fact that \replaced[id=bt]{WMHs do not generally manifest symmetrically across hemispheres}{the presence of WMHs violates that assumption},
we use the symmetric templates to compute the contralateral intensity
differences as an additional intensity feature.  For the FLAIR component, this contralateral
difference image is calculated from

@@

The segmentation probability images described above are used as feature images to provide a
spatial context for the random forest model prediction step.  Additional spatial contextual
feature images include the distance maps [@maurer2003] based on the csf, gray matter, and
deep gray matter images.  These latter images are intended to help distinguish white matter
hyperintensities from false positives induced by the partial voluming at the gray/white
matter interface.  A third set of images are based on the voxel location within the
space of the template.  The T1 image of the subject is registered to the T1 template
component using a B-spline variant [@Tustison:2013ac] of the well-known ANTs Symmetric
Normalization (SyN) algorithm [@Avants:2011ab].  Since the inverse transform is also
derived as part of the registration process, we can warp the voxel index locations
back to the space of the individual subject \replaced[id=nt]{which motivates similar work by others}{
Note that this is similar in motivation to the work
of} [@Anbeek:2004aa].  However, this previous work lacks the normalization to the standard
coordinate system provided by the template to dramatically improve spatial specificity
across all subjects.

@@

In previous brain tumor segmentation work [@Tustison:2015aa], it was demonstrated that a
concatenated supervised approach, whereby the prediction output from the first random forest
model serves as partial input for a second random forest model, can significantly improve
segmentation performance.  We do the same thing for the work described here \added[id=lw]{where we employ two stacked random forests (or two "stages")}.  The Stage 1
feature images of the training data (as described previously) are used to construct the
Stage 1 model.  The training data Stage 1 features are then used to produce the voxelwise
\added[id=lw]{"voting maps" (i.e., the classification count of each decision tree for each tissue label)} via the Stage 1 random forest model.  All the Stage 1 features plus the Stage 1 voting maps
are used as input to the Stage 2 model.  In addition, we use the Stage 1 voting maps as
tissue priors \added[id=lw]{(i.e., probabilistic estimates of the tissue spatial locations)} for a second application of the Atropos maximum aposteriori algorithm with an
additional Markov Random Field spatial prior (MAP-MRF) [@Avants:2011aa].  However, for
the second stage we use all three aligned preprocessed images for a multivariate
segmentation.  The resulting seven posterior probability images constitute a third additional feature image set for Stage 2.

@@

As pointed out in a recent comprehensive multiple sclerosis lesion segmentation review
[@Garcia-Lorenzo:2013aa], although the number of algorithms reported in the literature
is quite extensive, there were only four publicly available segmentation algorithms
at the time of writing \added[id=am]{this article}.  \replaced[id=dt]{In contrast to the current work,}{of which} none are based on supervised learning.  As we did for
our brain tumor segmentation algorithm [@Tustison:2015aa], all of the code described in
this work is publicly available through the open-source ANTs/ANTsR toolkits.  Through
ANTsR (an add-on toolkit which, in part, bridges ANTs and the R statistical project) we
use the \texttt{randomForest} package [@liaw2002] using the default settings with
2000 trees per model and 500 randomly selected samples per label per image.
\added[id=bt]{Note that we saw little variation in performance when these parameters were changed (i.e. up to 1000 random samples and as little as 1000 trees) which is consistent with our previous experience.}

In addition, similar to our previous offering [^1], we plan on creating
a self-encapsulated example to showcase the proposed methodology.  The fact that
the data will also be made available through the \replaced[id=lw]{Federal Interagency Traumatic Brain Injury Research (FITBIR)}{FITBIR} repository along
with the manual labelings will facilitate reproducibility on the part of the reader
as well as any interest in extending the proposed framework to other data sets.

@@

* Create Stage 1 feature images for all 24 subjects.
* For each of the 24 subjects:
    + sequester the current subject and corresponding feature images.
    + construct the Stage 1 random forest model from the remaining 23 subjects.
    + apply the Stage 1 random forest model to the feature images of the 23 training subjects.
    + the previous step produces the Stage 1 voting maps for all seven labels.
    + for each of the 23 subjects, perform a Bayesian-based segmentation with an MRF spatial
      prior using the seven voting maps are used as additional tissue priors.
    + construct the Stage 2 random forest model from all the Stage 1 feature images,
      seven voting maps, and seven posterior probability maps from the previous step.
    + send the sequestered subject through the random forest models for both stages.
    + compare the final results with the manually-defined white matter hyperintensity
      regions.

>>>>

MR images utilized for this initial report were acquired from a single scanner involved in the Chronic Effects of Neurotrauma Consortium’s (CENC) observational study (see Walker et al., this issue). Briefly, participants were Operation Iraqi Freedom/Operation Enduring Freedom (OIF/OEF) era Service Members and Veterans between the ages of 18-60 years with prior combat exposure and deployment(s). The feature images consisted of 26 subjects aged 39.6 ± 8.1 years (range 28 – 58 years). Within this cohort, 24 (92%) were considered positive for TBI based upon the potential concussive events (PCE) interview process described in detail in Walker et al., this issue). Each of the participants that were selected from the larger cohort of participants in this study demonstrated at least one white matter hyperintensity (but as many as 20) on FLAIR imaging.

@@

Crucial to these supervised segmentation approaches are the creation and selection of
"features" as input in conjunction with
expertly identified structures of interest for model construction.  For the targeted application in this work (i.e., WMHs),
tissue classification is performed at the voxelwise level.  In other words, each voxel
within the region of interest is sent through the ensemble of decision trees and receives
a set of classification votes from each tree thus permitting a regression or classification
solution.  Since this procedure is performed at the voxelwise level, intensity information
alone is insufficient for good segmentation performance due to the lack of spatial context.
For example, as pointed out in [@Neema:2009aa], higher intensities can be found at the
periventricular caps in normal subjects which often confounds automated lesion detection
algorithms.  Other potential confounds include MR signal inhomogeneity and noise.  Therefore,
even though machine learning and pattern recognition techniques are extremely powerful and
have significant potential, just as crucial to outcome is the creative construction and
deployment of salient feature images which we detail below.

@@

<!--

@@

-->

@@

<!--

![Representation of Stage 1 feature images for subject 01C1019.  The
FLAIR, T1-, and T2-weighted images are rigidly pre-aligned
[@Avants:2014aa] to the space of the T1 image.  The three modality images are then preprocessed
(N4 bias correction [@Tustison:2010ac] and adaptive denoising [@Manjon:2010aa]) followed by
application of standard ANTs brain extraction and $n$-tissue segmentation protocols using
the MMRR symmetric template and corresponding priors [@Tustison:2014ab] applied to the T1 image.
The feature images are then generated for voxelwise input to the RF model which results in
the voting maps illustrated on the right.  This gives a probabilistic classification of
tissue type.  Not shown are the probability and voting images for the brain stem and
cerebellum.](Figures/featureImages.png)

-->

As mentioned previously, input for each subject comprises FLAIR, T1-, and T2-weighted
acquisitions.  The FLAIR and T2 images are rigidly registered to the T1 image using
the open-source Advanced Normalization Tools (ANTs) [@Avants:2014aa].  The aligned images
are then preprocessed using the denoising algorithm of [@Manjon:2010aa] followed by
N4 bias correction [@Tustison:2010ac] which are then normalized to the intensity
range $[0,1]$.  Although we could have used an alternative intensity standardization
algorithm (e.g., [@nyul2000]), we found that a simple linear rescaling produced better results similar to previous work[@Tustison:2015aa].

<!--

@@

-->

The T1 image is then processed via the ANTs brain
extraction and normal tissue segmentation pipelines [@Tustison:2014ab].
Any effect of pathology is minimal because WMHs are not commonly seen in T1-weighted MRI.  This protocol involves preprocessing using N4 bias correction followed by a template-based strategy for brain extraction.  Once the
brain has been extracted, we apply a Bayesian-based segmentation algorithm
with a template-based prior probability strategy to segment the
parenchymal tissue types.  The result is a mask delineating the brain parenchyma and probabilistic estimates of the
CSF, gray matter, white matter, deep gray matter, brain stem, and
cerebellum.  These provide the expertly annotated labels for the first six tissue labels
given above.  The WMHs were manually identified by one of the authors
(J. R. S.) using the ITK-SNAP tool [@Yushkevich:2006aa].  Segmentation is performed using
the ANTs Atropos tool [@Avants:2011aa] and
multi-model optimal symmetric shape/intensity templates [@Tustison:2015aa] created from the public
MMRR data set  [@landman2011] (cf Figure 3).

<!--

@@

-->

To model the intensity information the first set of images simply includes the
preprocessed and normalized intensity FLAIR, T1, and T2 image voxel values.  We also calculate a set
of neighborhood statistics (mean, standard deviation, and skewness) feature images using
a Manhattan radius of one voxel given the typical size
of individual WMHs. For each of the normalized
images, we calculate the difference in intensities with the corresponding warped template
component.  Previous success in the international brain tumor segmentation
competition [@Menze:2015aa] was based on an important set of intensity features that were created
from multi-modal templates mentioned previously [@Tustison:2015aa] and listed in Table 1.  We employ the
same strategy here.  For example, the template
difference feature image for the FLAIR image, $S_{FLAIR}$ is calculated as:

@@

where

$$\phi_b: S  \leftrightarrow \underset{b}{\leftrightsquigarrow} T$$

is the transform
which maps from the individual subject space to the
template space and $T_{FLAIR}$ is the FLAIR template component.  Also,
to take advantage of the gross bilateral symmetry of the normal brain (in terms of both shape and intensity), and the fact that WMHs do not generally manifest symmetrically across hemispheres,
we use the symmetric templates to compute the contralateral intensity
differences as an additional intensity feature.  For the FLAIR component, this contralateral
difference image is calculated from

@@

The segmentation probability images described above are used as feature images to provide a
spatial context for the random forest model prediction step.  Additional spatial contextual
feature images include the distance maps [@maurer2003] based on the csf, gray matter, and
deep gray matter images.  These latter images are intended to help distinguish white matter
hyperintensities from false positives induced by the partial voluming at the gray/white
matter interface.  A third set of images are based on the voxel location within the
space of the template.  The T1 image of the subject is registered to the T1 template
component using a B-spline variant [@Tustison:2013ac] of the well-known ANTs Symmetric
Normalization (SyN) algorithm [@Avants:2011ab].  Since the inverse transform is also
derived as part of the registration process, we can warp the voxel index locations
back to the space of the individual subject which motivates similar work by others [@Anbeek:2004aa].  However, this previous work lacks the normalization to the standard
coordinate system provided by the template to dramatically improve spatial specificity
across all subjects.

@@

In previous brain tumor segmentation work [@Tustison:2015aa], it was demonstrated that a
concatenated supervised approach, whereby the prediction output from the first random forest
model serves as partial input for a second random forest model, can significantly improve
segmentation performance.  We do the same thing for the work described here where we employ two stacked random forests (or two "stages").  The Stage 1
feature images of the training data (as described previously) are used to construct the
Stage 1 model.  The training data Stage 1 features are then used to produce the voxelwise
"voting maps" (i.e., the classification count of each decision tree for each tissue label) via the Stage 1 random forest model.  All the Stage 1 features plus the Stage 1 voting maps
are used as input to the Stage 2 model.  In addition, we use the Stage 1 voting maps as
tissue priors (i.e., probabilistic estimates of the tissue spatial locations) for a second application of the Atropos maximum aposteriori algorithm with an
additional Markov Random Field spatial prior (MAP-MRF) [@Avants:2011aa].  However, for
the second stage we use all three aligned preprocessed images for a multivariate
segmentation.  The resulting seven posterior probability images constitute a third additional feature image set for Stage 2.

@@

As pointed out in a recent comprehensive multiple sclerosis lesion segmentation review
[@Garcia-Lorenzo:2013aa], although the number of algorithms reported in the literature
is quite extensive, there were only four publicly available segmentation algorithms
at the time of writing this article.  In contrast to the current work, none are based on supervised learning.  As we did for
our brain tumor segmentation algorithm [@Tustison:2015aa], all of the code described in
this work is publicly available through the open-source ANTs/ANTsR toolkits.  Through
ANTsR (an add-on toolkit which, in part, bridges ANTs and the R statistical project) we
use the _randomForest_ package [@liaw2002] using the default settings with
2000 trees per model and 500 randomly selected samples per label per image.
Note that we saw little variation in performance when these parameters were changed (i.e. up to 1000 random samples and as little as 1000 trees) which is consistent with our previous experience.

In addition, similar to our previous offering [^1], we plan on creating
a self-encapsulated example to showcase the proposed methodology.  The fact that
the data will also be made available through the Federal Interagency Traumatic Brain Injury Research (FITBIR) repository along
with the manual labelings will facilitate reproducibility on the part of the reader
as well as any interest in extending the proposed framework to other data sets.

@@

* Create Stage 1 feature images for all 24 subjects.
* For each of the 24 subjects:
    + sequester the current subject and corresponding feature images.
    + construct the Stage 1 random forest model from the remaining 23 subjects.
    + apply the Stage 1 random forest model to the feature images of the 23 training subjects.
    + the previous step produces the Stage 1 voting maps for all seven labels.
    + for each of the 23 subjects, perform a Bayesian-based segmentation with an MRF spatial
      prior using the seven voting maps as additional tissue priors.
    + construct the Stage 2 random forest model from all the Stage 1 feature images,
      seven voting maps, and seven posterior probability maps from the previous step.
    + send the sequestered subject through the random forest models for both stages.
    + compare the final results with the manually-defined white matter hyperintensity
      regions.
